On the relation of the entanglement spectrum to the Zak-phase

### 2020-06-18 draft.tex
Updated Restas formula and inhomogeneous plots

The polarization can be obtained as the many-body average of the position operator, but the latter has to be regularized for periodic boundary conditions \cite{Resta1997}. 
We can define a position operator for PBC as
\begin{equation}
\hat{X}_{\rm PBC} = e^{-i\frac{2\pi}{L}\hat{X}},
\end{equation}
where $\hat{X}$ is the sum of all the single-particle position operators.

We can now consider then the many-body average of the position operator over the ground state, defined as
\begin{align}
\expval{\hat{X}_{\rm PBC}}_0 \equiv -\frac{L}{2\pi} \rm{ Im }\ln\bra{\Psi_0}\hat{X}_{\rm PBC}\ket{\Psi_0},
\end{align}
%\carlos{This is the notation used in the Resta paper, but I'm not sure if it's confusing.}
where $\ket{\Psi_0}$ is the many-body ground state. 
Using that the ground state is a Slater determinant of the occupied single-particle eigenstates, $\ket{u^{p\mu}}$, it can be rewritten as~\cite{Resta1997}
\begin{align}
\expval{\hat{X}_{\rm PBC}}_0 = -\frac{L}{2\pi} \rm{Im }\ln \, {\rm det }\!' \, S,
\end{align}
where the matrix $S$ is given by
\begin{equation}
S_{p\mu,q\nu} = \sum_{j\alpha} u^{p\mu\, \ast}_{j \alpha} e^{-i\frac{2\pi}{L}j}u^{q\nu}_{j \alpha}
\end{equation}
and  ${\rm det }\!' $ indicates that the determinant is restricted to the space of \emph{occupied} single-particle states. 
The Zak phase can then be obtained as
\begin{equation}
\gamma = \expval{\hat{X}_{\rm PBC}}_0\frac{2\pi}{L} + \frac{1}{2}.
\end{equation}
\carlos{There is an extra 1/2 factor compared to Restas's formula which I don't know where it is coming from.}

@@

We have shown two cases were using expression \ref{eq:inh_simple} leads to the correct result, but this is in fact not the general case, as we can see in Fig.\ref{fig:simple_inh}.f . The average of a circular quantity is not uniquely defined and the choice we made in Eq.\ref{eq:inh_simple}, which equates to averaging the values in $(0,1)$ around $1/2$ (\carlos{is this clear?}) turns out to work in certain cases. How to perform this circular average is a priory not known, we can only judge after we compute the Zak phase independetly if we used the correct choice or not. Therefore in order to proceed we need to get rid of the circular average which leaves us with
\begin{align}
\bar{\chi} =\left(\frac{1}{L}\sum_{s=1}^L \chi_s \right)\, \rm{mod} \, 1,
\label{eq:inh_fixed}
\end{align}
but this however has to be done carefully. So far we have naively separated our eigenstates into \emph{left} and \emph{rigth} sectors, assigned the charges acordingly, and took the modulo 1 to eliminate the contribution of the bulk modes that localize at the center of A. The problem resides in the classification of these bulk modes. Even if the rest of the EOS remains constant, due to the inhomogeneity a bulk mode can cross the middle of region A, as shown in Fig.\ref{fig:loc_fix}, which will change $\chi$ by $1/L$, even though choosing $L/4$ as the separation between the \emph{left} and \emph{rigth} sectors is not physical. In order for \ref{eq:inh_fixed} to work properly the contribution from the bulk modes, which should vanish in the end, must remain constant across the different cuts. In order to achieve this we have to choose a cut-dependent threshold that the bulk modes do not cross. We choose this threshold, $l_s$, as the mean position of two of the eigenstates close to this middle point of A, as shown in Fig.\ref{fig:loc_fix}. This is done in practice using the Wannier states (\carlos{Do I explain exactly how do I do it numerically?}).
Note that in the thermodynamic limit it does not matter if the position of these eigenstates cross as they are degenerate, the only important thing is that the number of bulk modes at each sector remains constant (\carlos{Not sure about this phrasing either}).

@@

\begin{figure}[h!]
\centering
\includegraphics[width=70mm]{loc_fix.pdf}
\caption{Average position of the Wannier states in red and, slightly shifted, the localized entanglement eigenstates 
in blue. In black we show the threshold $l_s$ obtained as the mean value of the two Wannier states close de $19$ and $20$) $L=80,\lambda=10^{-6}, t = 1, t'= 2, m=0$}
\label{fig:loc_fix}
\end{figure}

As the sign assigned to the bulk modes becomes relevant in the inhomogeneous case one might think that it is better to use the open chain, $\chi_s = \Tr C_A^{{\rm open},s}$, as one avoids the issue of having to assign any signs at all. Although this is true we find another problem. By adiabatically tuning down the coupling between the $L$ and $1$ sites we see that the eigenvalues of the $left$ modes flow to either $0$ or $1$. This can lead to discontinuities in the number of bulk modes at $\xi=1$ which again result in an incorrect cut-average. \carlos{Do we want to include is? Or leave it for later?}

@@

However, as we did for systems without translational invariance, one can extend this identity for systems with pairing terms as well. Since one computes the Zak phase from the single-particle Bogoliubov-de Gennes Hamiltonian, we can obtain the correspondent ES and everything discussed above should apply as well. Take for example the kitaev chain, with the following Hamiltonian
\begin{align*}
H =& -\mu \sum_{i} c_i^\dagger c_i - t \sum_{i}\left( c_{i+1}^\dagger c_i + {\rm h.c.}\right) \\
&+  \sum_{i}\left( \Delta c_{i} c_{i+1} + {\rm h.c.}\right).
\end{align*}
Using the spinor $\phi_i^\dagger = (c_i^\dagger, c_i)$, we can rewrite the Hamiltonian as
\begin{align*}
&H = \frac{1}{2}\sum_i \phi^\dagger_i H_{BdG,ij} \phi_j,\\
&H_{BdG,ij} = -\mu \tau_z \delta_{ij} - (t \tau_z + i\Delta \tau_y )\delta_{i,j+1}- (t \tau_z - i\Delta \tau_y)\delta_{i,j-1}.
\end{align*}
Note that by taking $t' = \kappa = \kappa = 0$ in the BDI model, Eq. \ref{bdi_model}, we obtain
\begin{align*}
H_{ij} =& m \sigma_x\delta_{ij} + \frac{1}{2} t \left[(\sigma_x + i \sigma_y)\delta_{i,j+1} + (\sigma_x - i \sigma_y) \delta_{i,j-1} \right],
\end{align*}
which is just the Bogoliubov-de Gennes Hamiltonian of the Kitaev chain by rotating $x$ into $z$ and identifying $\mu = -m, t_{\rm Kit} = -t_{BDI}/2, \Delta = -t_{BDI}/2 $. From Fig.\ref{bdi_phase_diagram} we see that the $BDI$ model transitions from $\nu = 1$ to the trivial phase at $t=0, m=\pm t_{BDI}$, or alternatively, when $\mu = \pm 2 t_{Kit}$, which is the known result for the Kitaev chain. Therefore the results already shown for the BDI model studied here are also applicable to the Kitaev chain.

@@

>>>>

The polarization can be obtained from the many-body average of the position operator of electrons and ions, but the position operator has to be regularized for periodic boundary conditions \cite{Resta1997}. 
We can define a position operator for PBC as
\begin{equation}
\hat{X}_{\rm PBC} = e^{-i\frac{2\pi}{L}\hat{X}},
\end{equation}
where $\hat{X}$ is the sum of all the single-particle position operators.

We can now consider then the many-body average of the position operator over the ground state, defined as
\begin{align}
\expval{\hat{X}_{\rm PBC}}_0 \equiv -\frac{L}{2\pi} \rm{ Im }\ln\bra{\Psi_0}\hat{X}_{\rm PBC}\ket{\Psi_0},
\end{align}
%\carlos{This is the notation used in the Resta paper, but I'm not sure if it's confusing.}
where $\ket{\Psi_0}$ is the many-body ground state. 
Using that the ground state is a Slater determinant of the occupied single-particle eigenstates, $\ket{u^{p\mu}}$, it can be rewritten as~\cite{Resta1997}
\begin{align}
\expval{\hat{X}_{\rm PBC}}_0 = -\frac{L}{2\pi} \rm{Im }\ln \, {\rm det }\!' \, S,
\end{align}
where the matrix $S$ is given by
\begin{equation}
S_{p\mu,q\nu} = \sum_{j\alpha} u^{p\mu\, \ast}_{j \alpha} e^{-i\frac{2\pi}{L}j}u^{q\nu}_{j \alpha}
\end{equation}
and  ${\rm det }\!' $ indicates that the determinant is restricted to the space of \emph{occupied} single-particle states. Note that for an open chain the many-body average position of the ions will sit at a site when the chain length is odd and at a link when the length is even. The same result applies in the system with periodic boundary conditions, i.e. $\expval{\hat{X}_{PBC,+}}_{0}=(1+(-1)^{L})/4$.
The Zak phase can then be obtained as
\begin{align}
\gamma =& \left( \expval{\hat{X}_{\rm PBC}}_0-\expval{\hat{X}_{\rm PBC,+}}_0 \right)\frac{2\pi}{L} \\
=& \expval{\hat{X}_{\rm PBC}}_0\frac{2\pi}{L} -(1+(-1)^{L})/4.
\end{align}

@@

We have shown two cases were using expression \ref{eq:inh_simple} leads to the correct result, but this is in fact not the general case, as we can see in Fig.\ref{fig:simple_inh}.f . The average of a circular quantity is not uniquely defined and the choice we made in Eq.\ref{eq:inh_simple}, which equates to averaging the values in $(0,1)$ around $1/2$ (\carlos{is this clear?}) turns out to work in certain cases. How to perform this circular average is a priory not known, we can only judge after we compute the Zak phase independetly if we used the correct choice or not. Therefore in order to proceed we need to get rid of the circular average which leaves us with
\begin{align}
\bar{\chi} =\left(\frac{1}{L}\sum_{s=1}^L \chi_s \right)\, \rm{mod} \, 1,
\label{eq:inh_fixed}
\end{align}
but this however has to be done carefully. So far we have naively separated our eigenstates into \emph{left} and \emph{rigth} sectors, assigned the charges acordingly, and took the modulo 1 to eliminate the contribution of the bulk modes that localize at the center of A. The problem resides in the classification of these bulk modes. Even if the rest of the EOS remains constant, due to the inhomogeneity a bulk mode can cross the middle of region A, as shown in Fig.\ref{fig:wan1}.c, which will change $\chi$ by $1/L$, even though choosing $L/4$ as the separation between the \emph{left} and \emph{rigth} sectors is not physical. In order for Eq.\ref{eq:inh_fixed} to work properly the contribution from the bulk modes, which should vanish in the end, must remain constant across the different cuts. In order to achieve this we have to choose a cut-dependent threshold such that it is not crossed by the bulk modes, as shown in figure \ref{fig:wan1}.c. This is done in practice using the Wannier states.
Note that in the thermodynamic limit it does not matter if the position of these eigenstates cross as they are degenerate, the only important thing is that the number of bulk modes at each sector remains constant.

@@

\begin{figure}[h!]
\centering
\makebox[0pt]{
\subfloat[$\kappa_i$]{
  \includegraphics[width=35mm]{wan2_kap.pdf}
}
\subfloat[EOS (black) and $\chi_s$ mod $1$ (green)]{\includegraphics[width=35mm]{wan2_EOS.pdf}
}
}\hspace{0mm}

\makebox[0pt]{
\subfloat[Average position of the EOS eigenstates (blue) and Wannier states (red) around the threshold that separates the left and right sector (black).]{
  \includegraphics[width=35mm]{wan2_wan.pdf}
}
\subfloat[$\chi_s$ obtained using the appropiate left-right threshold (black), Zak phase (red) and $\bar{\chi}$ (dashed blue).]{\includegraphics[width=35mm]{wan2_essum.pdf}
}
}
\caption{Parameters used are: $t=1,t'=-2,m=0,\kappa'=0$. With the profile of $\kappa_i$ shown in (a) the width of $\chi_s$ is larger than one and therefore Eq.\ref{eq:inh_fixed} cannot be employed.}
\label{fig:wan1}
\end{figure}

\begin{figure}[h!]
\centering
\makebox[0pt]{
\subfloat[$\kappa_i$]{
  \includegraphics[width=35mm]{wan_kap.pdf}
}
\subfloat[EOS (black) and $\chi_s$ mod $1$ (green)]{\includegraphics[width=35mm]{wan_EOS.pdf}
}
}\hspace{0mm}

\makebox[0pt]{
\subfloat[Average position of the EOS eigenstates (blue) and Wannier states (red) around the threshold that separates the left and right sector (black).]{
  \includegraphics[width=35mm]{wan_wan.pdf}
}
\subfloat[$\chi_s$ obtained using the appropiate left-right threshold (black), Zak phase (red) and $\bar{\chi}$ (dashed blue).]{\includegraphics[width=35mm]{wan_essum.pdf}
}
}
\caption{Parameters used are: $t=1,t'=-2,m=0,\kappa'=0$. Adding disorder to an oscillating profile similar to the one in Fig.\ref{fig:wan1} shows that in general there is not enough information in $\chi_s$ mod $1$ to recover the Zak phase. }
\label{fig:wan2}
\end{figure}

In Fig.\ref{fig:wan1} we have shown an example where an oscillating profile for $\kappa_i$ results in a $\chi_s$ with a width which is larger than one, meaning that using Eq.\ref{eq:inh_fixed} is not posible in any case. Since we have chosen a continuous profile one might argue that one can recover the correct contribution of $\chi_s$, shown in Fig.\ref{fig:wan1}.c from $\chi_s$ mod $1$, there is however not possible a priori to know if that is the correct result. Furthermore one cannot do this in general, as shown in figure \ref{fig:wan2}. In this example we have added disorder to a similar oscillating profile for $\kappa_i$ that there are cases in general where clearly there is not enough information in $\chi_s$ mod $1$ to recover the Zak phase. \carlos{The profile in Fig.\ref{fig:wan2} is a really good example where there is no way one can recover the Zak phase from $\chi_s$ mod $1$. However I think the plot of average positions is too messy to understand what I mean by the bulk modes crossing $L/4$ and understand how the threshold is chosen, that's why I added the example in Fig.\ref{fig:wan1} but I dont know if we want to keep both.}

As the sign assigned to the bulk modes becomes relevant in the inhomogeneous case and leads to problems one might think that it is better to use the open chain, $\chi_s = \Tr C_A^{{\rm open},s}$, as one avoids the issue of having to assign any signs at all. Although this is true we find another issue. By adiabatically tuning down the coupling between the $L$ and $1$ sites we see that the eigenvalues of the $left$ modes flow to either $0$ or $1$. This can lead to discontinuities in the number of bulk modes at $\xi=1$ which again result in an incorrect cut-average. \carlos{Do we want to include is?}

@@

\carlos{I added this thinking it was generalizing on what Zaletel did but that was wrong, they also allow for interactions. Therefore I think it's too trivial for a section. It might be enough to mention that the above is valid for any quadratic Hamiltonian.}
However, as we did for systems without translational invariance, one can extend this identity for systems with pairing terms as well. Since one computes the Zak phase from the single-particle Bogoliubov-de Gennes Hamiltonian, we can obtain the correspondent ES and everything discussed above should apply as well. Take for example the kitaev chain, with the following Hamiltonian
\begin{align*}
H =& -\mu \sum_{i} c_i^\dagger c_i - t \sum_{i}\left( c_{i+1}^\dagger c_i + {\rm h.c.}\right) \\
&+  \sum_{i}\left( \Delta c_{i} c_{i+1} + {\rm h.c.}\right).
\end{align*}
Using the spinor $\phi_i^\dagger = (c_i^\dagger, c_i)$, we can rewrite the Hamiltonian as
\begin{align*}
&H = \frac{1}{2}\sum_i \phi^\dagger_i H_{BdG,ij} \phi_j,\\
&H_{BdG,ij} = -\mu \tau_z \delta_{ij} - (t \tau_z + i\Delta \tau_y )\delta_{i,j+1}- (t \tau_z - i\Delta \tau_y)\delta_{i,j-1}.
\end{align*}
Note that by taking $t' = \kappa = \kappa = 0$ in the BDI model, Eq. \ref{bdi_model}, we obtain
\begin{align*}
H_{ij} =& m \sigma_x\delta_{ij} + \frac{1}{2} t \left[(\sigma_x + i \sigma_y)\delta_{i,j+1} + (\sigma_x - i \sigma_y) \delta_{i,j-1} \right],
\end{align*}
which is just the Bogoliubov-de Gennes Hamiltonian of the Kitaev chain by rotating $x$ into $z$ and identifying $\mu = -m, t_{\rm Kit} = -t_{BDI}/2, \Delta = -t_{BDI}/2 $. From Fig.\ref{bdi_phase_diagram} we see that the $BDI$ model transitions from $\nu = 1$ to the trivial phase at $t=0, m=\pm t_{BDI}$, or alternatively, when $\mu = \pm 2 t_{Kit}$, which is the known result for the Kitaev chain. Therefore the results already shown for the BDI model studied here are also applicable to the Kitaev chain.

@@

\begin{figure*}[h!]
\centering
\makebox[0pt]{
\subfloat[${\rm Max}\bra{\psi_i}\sigma_x\ket{\psi_i}$ for $\xi_i >1/2$.]{
  \includegraphics[width=55mm]{bigsystemloc.pdf}
}
\subfloat[${\rm Log}_{10}(\chi - \gamma)$ ]{\includegraphics[width=55mm]{bigsystemerr.pdf}
}
}
\caption{$t = 1, t'= 2, m=0, \nu=2$. }
\label{lambdainh}
\end{figure*}

\begin{figure*}[h!]
\centering
\makebox[0pt]{
\subfloat[ ${\rm Max}\bra{\psi_i}\sigma_x\ket{\psi_i}$ for $\xi_i >1/2$]{
  \includegraphics[width=55mm]{smallsystemloc.pdf}
}
\subfloat[${\rm Log}_{10}(\chi - \gamma)$.]{\includegraphics[width=55mm]{smallsystemerr.pdf}
}
}
\caption{$t = 1, t'= 2, m=0, \nu=2$. }
\label{lambdainh}
\end{figure*}

\begin{figure*}[h!]
\centering
\makebox[0pt]{
\subfloat[${\rm Max}\bra{\psi_i}\sigma_x\ket{\psi_i}$ for $\xi_i >1/2$. ]{
  \includegraphics[width=55mm]{lambdaloc.pdf}
}
\subfloat[spectrum of $H_{ent}$]{\includegraphics[width=55mm]{lambdalocES.pdf}
}
}
\caption{$t = 1, t'= 2, m=0, \nu=2$. }
\label{lambdainh}
\end{figure*}

\begin{figure*}[h!]
\centering
\makebox[0pt]{
\subfloat[$\gamma$ (red) and $\chi$ (blue) ]{
  \includegraphics[width=55mm]{nonquanterror2.pdf}
}
\subfloat[${\rm Log}_{10}(\chi - \gamma)$]{\includegraphics[width=55mm]{nonquanterror3.pdf}
}
}
\caption{$t = 1, t'= 2, m=2, \kappa = 0.2 $. }
\label{lambdainh}
\end{figure*}

\section{Resta's formula}

Consider a 1D chain with L unit cells. For a half-filled two-orbital system the many-body ground-state is given by
\begin{equation}
\Psi_0(\vec{x}) = A\prod_{\mu} \psi_\mu (x_\mu),
\end{equation}
where $A$ is an antisymmetrizer and $\mu$ is restricted to the occupied states. The polarization is give by the expectation value of the operator
\begin{equation}
\hat{P}=\frac{-e}{L}\left(\sum_j X_j - \sum_j  x_j\right),
\end{equation}
where $X_j$ stands for the position operator of the nuclei, and $x_j$ the one for the electrons. Assuming a tight-binding approximation $X_j$ corresponds simply to the lattice positions.

Given periodic boundary conditions the polarization is obtained as
\begin{align*}
P =& \frac{-e}{2\pi}{\rm Im}\,{\rm Ln}\bra{\Psi_0}e^{i\frac{2\pi}{L} \left(\sum_j X_j - \sum_j  x_j\right)}\ket{\Psi_0} \\
=&\frac{-e}{2\pi}{\rm Im}\,{\rm Ln}\bra{\Psi_0}e^{-i\frac{2\pi}{L} \sum_j  x_j}\ket{\Psi_0}+\frac{-e(1+L)}{2}.
\end{align*}
Note that it is very important to include the contribution from the nuclei whenever we are working with periodic boundary conditions, as a new term dependent on system size appears. This can now be expressed as
\begin{align*}
P =& \frac{-e}{2\pi}{\rm Im}\,{\rm Ln} \, {\rm det}[S]+\frac{-e(1+L)}{2},
\end{align*}
where
\begin{equation}
S_{\mu \nu} = \int_0^L dx \, \psi^*_{\mu}(x)e^{i\frac{2\pi}{L}x}\psi_\nu(x).
\end{equation}
Consider now an spetial case where the system is translational invariant with only one occupied band. The indeces $\mu,\nu$ are now given by the momentum on the occupied band and we have
\begin{equation}
S_{q,q+1} = \int_0^L dx \, \psi^*_{q}(x)e^{i\frac{2\pi}{L}x}\psi_{q+1}(x),
\end{equation}
since all other elements of $S$ will vanish due to orthogonality proeprties of the Bloch functions. Now the determinant can be obtained as
\begin{equation}
{\rm det[S]} = \varepsilon_{2,3,...L,1}\prod_q S(q,q+1)
\end{equation}
Permuting two by two the $1$ back to the first position gives for the antisymmetric tensor $+1$ for odd $L$ and $-1$ for even $L$. The polarization is given therefore by
\begin{align*}
P =& \frac{-e}{2\pi}{\rm Im}\,{\rm Ln} \, \left\{e^{-i \pi (L+1)}\prod_q S(q,q+1)\right\}+\frac{-e(1+L)}{2} \\
=& \frac{-e}{2\pi}{\rm Im}\,{\rm Ln} \, \left\{\prod_q S(q,q+1)\right\}
\end{align*} 
which coincides with the known result for the discretized Zak phase. Both the term due to the nuclei and this sign coming from the antisymmetric tensor are missing from the Resta paper, but since they cancel each other its final result was ok. It's only when you compute things without translational invariance that the issue arised.
